Binning Strategies and Related Loss for Large Data

### 2015-05-12 Paper.Rnw
done with the conclusions

Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation can come at the cost of losing precision in the spatial and frequency information.

We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms. The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution. However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid creating artificial stripes if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.  It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.

Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure. It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.

>>>>

Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.

We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms. The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution. However, standard binning algorithm is superior due to lower spatial loss. For data with resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.  We were also able to show through simulation that  a reasonable default for the binning origin is an offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well even for skewed data.

Spatial binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure. It is left to the plot designer to decide how much spatial information they are willing to sacrifice in order to simplify the display of density structure. %; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss.
